%This document was written by using TeXShop Version 1 (v1.43) for Mac
\documentclass[12pt, english]{article}
\usepackage[T1]{fontenc}
\usepackage[applemac]{inputenc}
\usepackage[ngerman]{babel} 
\usepackage{geometry}
\usepackage{amsmath} 
\usepackage{amsthm} 
\usepackage{amssymb}
\geometry{a4paper}
\usepackage{graphicx}
%\usepackage{epstopdf}
\usepackage{setspace}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `basename #1 .tif`.png}

\begin{document}

\begin{center}
\thispagestyle{empty}
{\LARGE \bf 
 Max-Planck-Institute for Meteorology \\
 \LARGE in Hamburg \\ [16ex]
\LARGE $\mathbf{tsplot}$ \\ [1ex]
\Large a tool to plot time series data \\

\Large Documentation \\
[18ex] 
\Large Frank Ršske \\
frank.roeske@zmaw.de \\
August, 2007 \\
modified by Helmuth Haak\\
November 2007\\
}
\end{center}
\newpage

\onehalfspacing
%\addtocounter{page}{-1}

%\tableofcontents

%\section{Description of $\mathbf{tsplot}$}
{
%$\mathbf{tsplot}$ supports one as well as two hyphen options.
%Both, {\tt tsplot $--$help} and {\tt tsplot $-$h} produce the same following help text:\\

{\tt
\noindent $\mathbf{NAME}$ \\
tsplot - To plot time series \\
 \\
$\mathbf{SYNOPSIS}$ \\
$\mathbf{tsplot}$ [\underline{OPTIONS}] [\underline{FILE(S)}] \\
 \\
$\mathbf{DESCRIPTION}$ \\
$\mathbf{tsplot}$ generates xy-plots in \underline{PostScript} from MPIOM data by using GMT tools. \\
 \\
\underline{OPTIONS} \\
--help,{~ }-h         Print this text. \\
--codes,  -c \underline{CODES}  Codes: Commas for lists and hyphens for ranges. \\
\indent \indent \indent \indent \indent \indent No overlap. (Default: '1') \\
--resol,  -r \underline{RESOL}  Time resolution: 0 for yearly, 1 for monthly \\
\indent \indent \indent \indent \indent \indent and 2 for daily. (Default: '1') \\
\indent \indent \indent \indent \indent  \indent It is dependent on what the FILE(S) allow(s) for. \\
--nplots,-n \underline{NPLOTS}  Number of plots per page  (Default: '2') \\
--xticks,-x \underline{XTICKS}  Tickmarks on the x-axis   (Default: '10') \\
--yticks,-y \underline{YTICKS}  Tickmarks on the y-axis   (Default: '10') \\
--suppr, -s         Suppress plotting variable names. \\
--gaps,{~ }-g         Look for gaps. \\
--vfile, -v \underline{FILE}    File of variable names    \\
\indent \indent \indent \indent \indent  \indent (Default: 'TSVAR' is created from mpiom.partab) \\
--dfile, -d \underline{FILE}    File of description texts \\
\indent \indent \indent \indent \indent  \indent (Default: 'TSDESC' is created from mpiom.partab) \\
--ufile, -u \underline{FILE}    File of physical units \\
 \indent \indent \indent \indent \indent  \indent  (Default: 'TSUNIT' is created from mpiom.partab) \\
\underline{FILE(S)} \indent \indent \indent Up to ten file(s) of time series data written in the \\
\indent \indent \indent \indent \indent{~ }EXTRA format. Names must include just one '\_'. \\
\indent \indent \indent \indent \indent{~ }(Default: 'ZEITSER.ext\_'). \\
\indent \indent \indent \indent \indent{~ }Characters after '\_' are interpreted as experiment names. \\
\indent \indent \indent \indent \indent{~ }Experiment names may not contain '\_0', '\_1' or '\_2'. \\


\newpage

\noindent If \underline{OPTIONS} are not given, $\mathbf{tsplot}$ inquires them interactively \\
and shows the information extracted from the files. \\
At least one \underline{OPTION} is necessary to run non-interactively. \\
}

\noindent \underline{CODES} may be given by a list of numbers separated by commas. The
list entries may be single numbers or ranges of numbers indicated by
hyphens. e.g. {\tt 1-4,6,9-11}. The numbers specified that way must occur just
once and may not overlap.

\noindent \underline{RESOL} may be 0 for yearly, 1 for monthly, and 2 for daily
resolution. $\mathbf{tsplot}$ examines the \underline{FILE(S)} for their
resolutions. If they have daily resolution, \underline{RESOL} may be 0, 1, or
2. If they have monthly resolution, \underline{RESOL} may only be 0 or 1, and
if they have yearly resolution, \underline{RESOL} may only be 0. If more than
one file is given, that one with the most coarse resolution determines the
range allowed for \underline{RESOL}. Dependent on this structure,
$\mathbf{tsplot}$ decides whether files have to be averaged or not. If the
file resolution is equal to \underline{RESOL}, a symbolic link is created.

\noindent \underline{NPLOTS} may be any number greater than or equal to 1. It determines
the number of time series plots on one page. On such a page the plots are
arranged bottom-up.

\noindent \underline{XTICKS} and \underline{YTICKS} determine the number of tickmarks on
the x- and y-axis, respectively.  {However}, these numbers are only
approximations of those really plotted, because they are adjusted to match
reasonable intervals.

\noindent The {\tt$--$suppr} or {\tt$-$s} option allows to suppress plotting variable
names that might be interesting for publication purposes. Default is no.

\noindent The {\tt$--$gaps} or {\tt$-$g} option allows to search through the time series
files for gaps on a yearly base. If gaps are found, the time series are
decomposed into several subseries. They are plotted independently of each
other so that disturbing connection lines between those subseries are not
plotted. Default is no.

\noindent The {\tt$--$vfile} or {\tt$-$v}, {\tt$--$dfile} or {\tt$-$d}, {\tt$--$ufile}
or {\tt$-$d} options allow to specify three files containing texts needed to
label the time series plots.  For all options which were described defaults
are given.

\noindent $\mathbf{tsplot}$ distinguishes an interactive and a non-interactive mode.  If
\underline{OPTIONS} are not given, the script runs interactively. If at least
one option is supplied, it runs non- interactively.  This should be preferred
when the content of the file or the files is not known. However, the script
extracts information about the files which is shown only in the interactive
mode. These informations include the first and the last year and if a year is
not complete, the number of codes and the temporal resolution. If more than
one file of time series data is given, the script determines the minimum of
those numbers to account for future changes in the number of codes.
Furthermore, in the interactive mode the script informs about what it is doing
at the moment - similar to a verbose mode.

\noindent If \underline{FILE(S)} are not given, the script looks in the current
directory by default for files whose names begin with {\tt ZEITSER.ext\_}. If
\underline{FILE(S)} are given, their names must include just one underscore
sign ( \_ ), because the character string which follows this sign is
interpreted as experiment name.

\noindent {While} $\mathbf{tsplot}$ is running some intermediate files are created which
are removed again when the script finishes properly. The names of those files
consist of the original file name extended by a \_ and a number denoting the
resolution, followed by another \_ and a number denoting the code. In
addition, files are created whose names are furthermore extended by \_info,
\_info2, \_info3, and \_output. The files whose names end on \_info3 can
further on be extended by another \_ and the subseries number, if gaps were
encountered.

\noindent In order not to confuse those file names in regard to the temporal resolution,
the names of the \underline{FILE(S)} may not include \_0, \_1, or
\_2. Therefore, also the experiment names may not include these three
character strings.

\noindent Up to ten files are allowed. This somewhat arbitrary limit is linked to the
way colors are defined. Ten names for colors are used in this order: black,
red, gold, green, cyan, blue, purple, brown, grey, pink.  }


\newpage

\label{tsplot}
\noindent In $\mathbf{tsplot}$ five GMT tools are used,

\begin{tabular}{ll}
\\
gmtset&to set GMT variables \\
minmax&to determine minima and maxima of the data to be plotted \\
psbasemap&to draw axis' and labels \\
psxy&to plot the data \\
pstext&to plot MPIOM variables, physical units, experiment names. \\
\\
\end{tabular}

\noindent These commands produce the file {\tt plot.ps} containing PostScript code. This
file is controlled by the GMT options {\tt -O} and {\tt -K} which are
available for all these commands. Option {\tt -O} denotes the overlay mode,
and option {\tt -K} refers to as the append mode. That GMT command which is
invoked first, may use only the option {\tt -K}, that one which is invoked
last, may use only the option {\tt -O}, and all those invoked in between must
use both options.

\noindent $\mathbf{tsplot}$ has been tested on a Linux machine ('cross') and on a SunOS
machine ('yang'). Although the Bourne Again Shell (bash) was available on both
computers, this shell showed different behaviour in regard to reading options
and the {\tt tr} command. Under Linux the {\tt getopt} command is applied for
reading options but under SunOS the {\tt getopts} command is used instead. The
{\tt tr} command is applied in a multiple way to achieve similar behaviour on
both machines. Also, the GNU {\tt awk} ({\tt gawk}) command is used instead of
{\tt awk} to provide similar behaviour. {\tt gawk} is needed to calculate real
numbers, whereas calculating integers is done by the {\tt expr} command.  On
the Linux machine the GMT version 4.1.4 is applied, whereas on the SunOS
machine GMT version 4.2 is used instead of version 4.1.3 which showed a
different behaviour. If $\mathbf{tsplot}$ is ported to a machine other than
Linux or SunOS, the paths for CDO and GMT have to be changed accordingly and
the shell commands controlled by the {\tt uname} shell variable should be
traced carefully.

\newpage
\section{Timeseries in MPIOM}
{
\label{changes}
\noindent The time series data from MPIOM consist of 151 codes (Summer 2007). These are
 written {\tt formatted} (TIMESER.asc) and/or {\tt unformatted} (TIMESER.ext)
 by using the EXTRA format.  A record written in EXTRA format can be read in a
 FORTRAN program by
\newline \\ {\tt \indent READ(10) IDATE,ICODE,ILEVEL,NSIZE \\ \indent READ(10)
 (FIELD(ISIZE),ISIZE=1,NSIZE) \\ } \\ whereas {\tt IDATE} denotes the date,
 {\tt ICODE} the code, {\tt ILEVEL} the level, and {\tt NSIZE} the size of the
 record (http://www.mad.zmaw.de/Pingo/downloads.html).\\

\noindent PSEUDO EXTRA : If the user wishes all 151 codes can be written at once in one record. This
 however violates the principle of the EXTRA format, because one record should
 be assigned to just one code (see {\tt ICODE}). Therefore, this format is
 called PSEUDO EXTRA format. For $\mathbf{tsplot}$ to be able to plot time
 series, the data have to be transposed. Unfortunately, CDO does not support a
 transpose function. The {\tt OCECTL} Namelist from MPIOM includes a variable {\tt ltstranspose} of
 type {\tt logical}. If it is {\tt .false.} the PSEUDO EXTRA format is applied 
 and if it is {\tt .true.}, the EXTRA format is applied so that only one code is
  written in one record, i.e. {\tt NSIZE=1}.\\

\noindent The namelist variable {\tt itsdiag} controls the output. Eight settings are possible.

\begin{tabular}{rl}
\\
0&no output \\
1&one snapshot per day \\
2&monthly averaged snapshot \\
3&yearly averaged snapshot \\
4&output every timestep \\
5&daily average \\
6&monthly mean of daily averages \\
7&yearly mean of daily averages \\
\end{tabular}
\newline
 
\noindent In {\tt mpiom.f90} the variable {\tt itsdiag} controls how often the
subroutine {\tt diagnosis} is called. This subroutine is called each time
step, if {\tt itsdiag} is greater than are equal to 4. Otherwise {\tt
diagnosis} is called once per day as in the former version.  In {\tt
mo\_diagnosis.f90} the new subroutine {\tt write\_timeseries} is called, if
{\tt itsdiag} is greater than or equal to 1. In {\tt write\_timeseries} the
variable {\tt itsdiag} controls the accumulating and averaging of time series
data as described above.  In addition, three new files are generated by {\tt
write\_timeseries} containing texts used by $\mathbf{tsplot}$ to label the
time series plots. File {\tt TSVAR} contains the variable names as used in
MPIOM. File {\tt TSDESC} includes the scientific expressions of these
variables to be used as titles for the plots. File {\tt TSUNIT} provides the
physical units to label the y-axis of the plots (see Appendix).  These files
are written only when {\tt diagnosis} is called the first time. 

\section*{Appendix}
\noindent Examples:\\
{\tt
As first step the user needs to cat the individual time series in to one file.\\
 
cat \$WRKSHR//TIMESER.????0101\_????1231.ext >> TIMESER\_hel9994 \\

\noindent As second step make the plot, e.g. for codes 1-16 and 32-151\\

tsplot -c1-16,32-151 -r 0 -n 2 TIMESER\_hel9994 \\
}
\newpage

%\noindent TIMESER VARIABLES: \\
\begin{tabular}{rlll}
  1  &  PSIGULF	 & max\_of\_barotropic\_streamfunction\_in\_subtropical\_atlantic	 & m3 s-1\\
  2  &  PSIKURO	 & max\_of\_barotropic\_streamfunction\_in\_subtropical\_pacific	 & m3 s-1\\
  3  &  PSIBANDA & barotropic\_transport\_through\_indonesian\_archipelago	 & m3 s-1\\
  4  &  PSIDRAKE & barotropic\_transport\_through\_drake\_passage	 & m3 s-1\\
  5  &  PSIBERING & barotropic\_transport\_through\_bering\_strait	 & m3 s-1\\
  6  &  PSISPG	 & max\_of\_barotropic\_streamfunction\_in\_subpolar\_atlantic	 & m3 s-1\\
  7  &  CO2	 & mass\_fraction\_of\_carbon\_dioxide\_in\_air	 & ppm\\
  8  &  CO2FLUX	 & downward\_carbon\_flux\_at\_surface	 & mole m-2\\
  9  &  AABW2	 & mass\_transport\_below\_1000m\_in\_atlantic\_around\_60N	 & m3 s-1\\
 10  &  NADW2	 & mass\_transport\_below\_1000m\_in\_atlantic\_around\_60N	 & m3 s-1\\
 11 &   AABW3	 & mass\_transport\_below\_1000m\_in\_atlantic\_around\_50N	 & m3 s-1\\
 12 &   NADW3	 & mass\_transport\_below\_1000m\_in\_atlantic\_around\_50N	 & m3 s-1\\
 13 &   AABW4	 & mass\_transport\_below\_1000m\_in\_atlantic\_around\_30N	 & m3 s-1\\
 14 &   NADW4	 & mass\_transport\_below\_1000m\_in\_atlantic\_around\_30N	 & m3 s-1\\
 15 &   AABW5	 & mass\_transport\_below\_1000m\_in\_atlantic\_around\_30S	 & m3 s-1\\
 16 &   NADW5	 & mass\_transport\_below\_1000m\_in\_atlantic\_around\_30S	 & m3 s-1\\
 17 &   TVQUER1	 & heat\_transport\_by\_advection\_in\_pacific\_at\_65N	 & W\\
 18 &   SVQUER1	 & salt\_transport\_by\_advection\_in\_pacific\_at\_65N	 & g s-1\\
 19 &   TMERCI1	 & mass\_transport\_in\_pacific\_at\_65N	 & m3 s-1\\
 20 &   TVQUER2	 & heat\_transport\_by\_advection\_in\_atlantic\_at\_60N	 & W\\
 21 &   SVQUER2	 & salt\_transport\_by\_advection\_in\_atlantic\_at\_60N	 & g s-1\\
 22 &   TMERCI2	 & mass\_transport\_in\_atlantic\_at\_60N	 & m3 s-1\\
 23 &   TVQUER3	 & heat\_transport\_by\_advection\_in\_atlantic\_at\_50N	 & W\\
 24 &   SVQUER3	 & salt\_transport\_by\_advection\_in\_atlantic\_at\_50N	 & g s-1\\
 25 &   TMERCI3	 & mass\_transport\_in\_atlantic\_at\_50N	 & m3 s-1\\
 26 &   TVQUER4	 & heat\_transport\_by\_advection\_in\_atlantic\_at\_30N	 & W\\
 27 &   SVQUER4	 & salt\_transport\_by\_advection\_in\_atlantic\_at\_30N	 & g s-1\\
 28 &   TMERCI4	 & mass\_transport\_in\_atlantic\_at\_30N	 & m3 s-1\\
 29 &   TVQUER5	 & heat\_transport\_by\_advection\_in\_atlantic\_at\_30S	 & W\\
 30 &   SVQUER5	 & salt\_transport\_by\_advection\_in\_atlantic\_at\_30S	 & g s-1\\
 31 &   TMERCI5	 & mass\_transport\_in\_atlantic\_at\_30S	 & m3 s-1\\
 32 &   TVNET2	 & net\_heat\_transport\_by\_advection\_in\_atlantic\_at\_60N	 & W\\
 33 &   SVNET2	 & net\_salt\_transport\_by\_advection\_in\_atlantic\_at\_60N	 & g s-1\\
 34 &   TVNET3	 & net\_heat\_transport\_by\_advection\_in\_atlantic\_at\_50N	 & W\\
 35 &   SVNET3	 & net\_salt\_transport\_by\_advection\_in\_atlantic\_at\_50N	 & g s-1\\
 36 &   TVNET4	 & net\_heat\_transport\_by\_advection\_in\_atlantic\_at\_30N	 & W\\
 37 &   SVNET4	 & net\_salt\_transport\_by\_advection\_in\_atlantic\_at\_30N	 & g s-1\\
 38 &   TVNET5	 & net\_heat\_transport\_by\_advection\_in\_atlantic\_at\_30S	 & W\\
 39 &   SVNET5	 & net\_salt\_transport\_by\_advection\_in\_atlantic\_at\_30S	 & g s-1\\
 40 &   TRBERING & mass\_transport\_through\_bering\_strait	 & m3 s-1\\
\end{tabular}
\newpage
\begin{tabular}{rlll}
 41 &   TRDENMARK & overflow\_transport\_through\_denmark\_strait	 & m3 s-1\\
 42 &   TRFAROER & overflow\_transport\_through\_faroer\_bank\_channel	 & m3 s-1\\
 43 &   SFRAM	& seaice\_transport\_through\_fram\_strait	 & m3 s-1\\
 44 &   ICEARE\_ARC	 & seaice\_area	 & m2\\
 45 &   ICEVOL\_ARC	 & seaice\_volume	 & m3\\
 46 &   HFL\_ARC	 & downward\_heatflux\_into\_ocean	 & W  \\
 47 &   WFL\_ARC	 & downward\_waterflux\_into\_ocean	 & m3 s-1  \\
 48 &   SST\_ARC	 & sea\_surface\_temperature	 & deg C  \\
 49 &   SSS\_ARC	 & sea\_surface\_salinity	 & psu  \\
 50 &   T200\_ARC	 & potential\_temperature	 & deg C  \\
 51 &   S200\_ARC	 & salinity	 & psu  \\
 52 &   T700\_ARC	 & potential\_temperature	 & deg C  \\
 53 &   S700\_ARC	 & salinity	 & psu  \\
 54 &   T2200\_ARC	 & potential\_temperature	 & deg C  \\
 55 &   S2200\_ARC	 & salinity	 & psu  \\
 56 &   ICEARE\_GIN	 & seaice\_area	 & m2\\
 57 &   ICEVOL\_GIN	 & seaice\_volume	 & m3\\
 58 &   HFL\_GIN	 & downward\_heatflux\_into\_ocean	 & W  \\
 59 &   WFL\_GIN	 & downward\_waterflux\_into\_ocean	 & m3 s-1  \\
 60 &   SST\_GIN	 & sea\_surface\_temperature	 & deg C  \\
 61 &   SSS\_GIN	 & sea\_surface\_salinity	 & psu  \\
 62 &   T200\_GIN	 & potential\_temperature	 & deg C  \\
 63 &   S200\_GIN	 & salinity	 & psu  \\
 64 &   T700\_GIN	 & potential\_temperature	 & deg C  \\
 65 &   S700\_GIN	 & salinity	 & psu  \\
 66 &   T2200\_GIN	 & potential\_temperature	 & deg C  \\
 67 &   S2200\_GIN	 & salinity	 & psu  \\
 68 &   ICEARE\_LAB	 & seaice\_area	 & m2\\
 69 &   ICEVOL\_LAB	 & seaice\_volume	 & m3\\
 70 &   HFL\_LAB	 & downward\_heatflux\_into\_ocean	 & W  \\
 71 &   WFL\_LAB	 & downward\_waterflux\_into\_ocean	 & m3 s-1  \\
 72 &   SST\_LAB	 & sea\_surface\_temperature	 & deg C \\ 
 73 &   SSS\_LAB	 & sea\_surface\_salinity	 & psu  \\
 74 &   T200\_LAB	 & potential\_temperature	 & deg C  \\
 75 &   S200\_LAB	 & salinity	 & psu  \\
 76 &   T700\_LAB	 & potential\_temperature	 & deg C  \\
 77 &   S700\_LAB	 & salinity	 & psu  \\
 78 &   T2200\_LAB	 & potential\_temperature	 & deg C  \\
 79 &   S2200\_LAB	 & salinity	 & psu  \\
 80 &   ICEARE\_NAT	 & seaice\_area	 & m2\\
\end{tabular}
\newpage
\begin{tabular}{rlll}
 81 &   ICEVOL\_NAT	 & seaice\_volume	 & m3\\
 82 &   HFL\_NAT	 & downward\_heatflux\_into\_ocean	 & W  \\
 83 &   WFL\_NAT	 & downward\_waterflux\_into\_ocean	 & m3 s-1  \\
 84 &   SST\_NAT	 & sea\_surface\_temperature	 & deg C  \\
 85 &     SSS\_NAT	 & sea\_surface\_salinity	 & psu  \\
 86 &    T200\_NAT	 & potential\_temperature	 & deg C  \\
 87 &   S200\_NAT	 & salinity\_at\_200m	 & psu  \\
 88 &   T700\_NAT	 & potential\_temperature	 & deg C  \\
 89 &   S700\_NAT	 & salinity	 & psu  \\
 90 &   T2200\_NAT	 & potential\_temperature	 & deg C  \\
 91 &   S2200\_NAT	 & salinity	 & psu  \\
 92 &   ICEARE\_ATL	 & seaice\_area	 & m2\\
 93 &   ICEVOL\_ATL	 & seaice\_volume	 & m3\\
 94 &   HFL\_ATL	 & downward\_heatflux\_into\_ocean	 & W  \\
 95 &   WFL\_ATL	 & downward\_waterflux\_into\_ocean	 & m3 s-1  \\
 96 &   SST\_ATL	 & sea\_surface\_temperature	 & deg C  \\
 97 &   SSS\_ATL	 & sea\_surface\_salinity	 & psu  \\
 98 &   T200\_ATL	 & potential\_temperature	 & deg C  \\
 99 &   S200\_ATL	 & salinity	 & psu  \\
100 &   T700\_ATL	 & potential\_temperature	 & deg C  \\
101  &  S700\_ATL	 & salinity	 & psu  \\
102  &  T2200\_ATL	 & potential\_temperature	 & deg C  \\
103 &   S2200\_ATL	 & salinity	 & psu  \\
104 &   ICEARE\_SO	 & seaice\_area	 & m2\\
105 &   ICEVOL\_SO	 & seaice\_volume	 & m3\\
106 &   HFL\_SO	 & downward\_heatflux\_into\_ocean	 & W  \\
107 &   WFL\_SO	 & downward\_waterflux\_into\_ocean	 & m3 s-1  \\
108 &   SST\_SO	 & sea\_surface\_temperature	 & deg C  \\
109 &   SSS\_SO	 & sea\_surface\_salinity	 & psu  \\
110 &   T200\_SO	 & potential\_temperature	 & deg C  \\
111  &    S200\_SO	 & salinity	 & psu  \\
112 &   T700\_SO	 & potential\_temperature	 & deg C  \\
113  &    S700\_SO	 & salinity	 & psu  \\
114 &   T2200\_SO	 & potential\_temperature	 & deg C  \\
115 &   S2200\_SO	 & salinity	 & psu  \\
116 &   ICEARE\_PAC	 & seaice\_area	 & m2\\
117 &   ICEVOL\_PAC	 & seaice\_volume	 & m3\\
118 &   HFL\_PAC	 & downward\_heatflux\_into\_ocean	 & W  \\
119 &   WFL\_PAC	 & downward\_waterflux\_into\_ocean	 & m3 s-1  \\
120 &   SST\_PAC	 & sea\_surface\_temperature	 & deg C  \\
\end{tabular}
\newpage
\begin{tabular}{rlll}
121 & SSS\_PAC & sea\_surface\_salinity	 & psu  \\
122 & T200\_PAC	 & potential\_temperature	 & deg C  \\
123 & S200\_PAC	 & salinity	 & psu  \\
124 & T700\_PAC	 & potential\_temperature	 & deg C  \\
125 & S700\_PAC	 & salinity	 & psu  \\
126 & T2200\_PAC & potential\_temperature	 & deg C  \\
127 & S2200\_PAC & salinity	 & psu  \\
128 & ICEARE\_NI3 & seaice\_area	 & m2\\
129 & ICEVOL\_NI3 & seaice\_volume	 & m3\\
130 & HFL\_NI3 & downward\_heatflux\_into\_ocean	 & W  \\
131 & WFL\_NI3 & downward\_waterflux\_into\_ocean	 & m3 s-1  \\
132 & SST\_NI3 & sea\_surface\_temperature	 & deg C  \\
133 & SSS\_NI3 & sea\_surface\_salinity	 & psu  \\
134 & T200\_NI3 & potential\_temperature	 & deg C  \\
135 & S200\_NI3 & salinity	 & psu  \\
136 & T700\_NI3 & potential\_temperature	 & deg C  \\
137 & S700\_NI3 & salinity	 & psu  \\
138 & T2200\_NI3 & potential\_temperature	 & deg C  \\
139 & S2200\_NI3 & salinity	 & psu  \\
140 & ICEARE\_GLO & seaice\_area	 & m2\\
141 & ICEVOL\_GLO & seaice\_volume	 & m3\\
142 & HFL\_GLO & downward\_heatflux\_into\_ocean	 & W  \\
143 & WFL\_GLO & downward\_waterflux\_into\_ocean	 & m3 s-1  \\
144 & SST\_GLO & sea\_surface\_temperature	 & deg C  \\
145 & SSS\_GLO & sea\_surface\_salinity	 & psu  \\
146 & T200\_GLO & potential\_temperature	 & deg C  \\
147 & S200\_GLO & salinity	 & psu  \\
148 & T700\_GLO & potential\_temperature	 & deg C  \\
149 & S700\_GLO & salinity	 & psu  \\
150 & T2200\_GLO & potential\_temperature	 & deg C  \\
151 & S2200\_GLO & salinity	 & psu  \\
\end{tabular}
}

\end{document}
